Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2021-11-01 Multilevel Missing Data/Multilevel Missing Data.Rmd
Outline paper incl modeling steps

---
documentclass: jss
author:
  - name: Hanne Oberman
    affiliation: Utrecht University
    address: |
      | Padualaan 14
      | 3584 CH Utrecht
    email: \email{h.i.oberman@uu.nl}
    url: https://hanneoberman.github.io/
  - name: Johanna Munoz Avila
    affiliation: University Medical Center Utrecht
  - name: Valentijn
    affiliation: University Medical Center Utrecht \AND
  - name: Gerko Vink
    affiliation: Utrecht University
  - name: Thomas Debray
    affiliation: University Medical Center Utrecht 
      # | Julius Center for Health Sciences and Primary Care, 
      # | University Medical Center Utrecht, Utrecht, The Netherlands
    #affiliation2:
title:
  formatted: "Imputation of Incomplete Multilevel Data with \\pkg{mice}"
  # If you use tex in the formatted title, also supply version without
  plain:     "Imputation of Incomplete Multilevel Data with mice"
  # For running headers, if needed
  short:     "\\pkg{mice}: Multilevel"
abstract: >
  Tutorial paper on imputing incomplete multilevel data with \pkg{mice}. Including methods for ignorable and non-ignorable missingness. AKA missing on so many levels.
keywords:
  # at least one keyword must be supplied
  formatted: [missing data, multilevel, clustering, "\\proglang{R}"]
  plain:     [missing data, multilevel, clustering, R]
preamble: >
  \usepackage{amsmath}
output: rticles::jss_article
bibliography: ../References/multilevelmice.bib
---

@@

- What is clustering/multilevel data? In this paper, we discuss grouped observations, not longitudinal data (within-patient clustering). What do we mean by clustering? In the medical field: Clustering by studies (IPDMA), hospitals in registries, multi-center studies etc. In other fields: e.g. official stats clustering at country-level, or social sciences clustering at school-level (related to the sampling design).

- What is heterogeniety (i.e. variability within studies vs. variability between studies)? ADD: Figure to show difference between patient-level datapoints vs cluster-level datapoints. And different data frame formats.

- What methods are required to analyze multilevel data?	Random effects for intercept term, predictor effects, and/or variance residual error. Add references, e.g. @hox17.

@@

- Why/where does missingness occur? I.e., not only patient-level but also cluster-level.

- How can we categorize this? Systematic vs sporadic missingness, see @resc13. ADD: visualization of systematic vs sporadic missingness. Within systematic we have always missing (same value per cluster) and non-measured variables (may differ per patient). TODO: adjust md pattern to match text.

```{r patterns, echo=FALSE}
# monotone vs non-monotone
patterns <- rbind(
  tibble(pat = rep("Monotone", 5), rownr = 5:1, X = rep(1, 5), Y = c(1,1,1,0, 0), Z = c(1, rep(0,4))) %>% pivot_longer(cols = c(X, Y, Z)),
  tibble(pat = rep("Non-monotone", 5), rownr = 5:1, X = c(0, rep(1,4)), Y = c(1,1,1,0, 0), Z = c(1, rep(0,3), 1)) %>% pivot_longer(cols = c(X, Y, Z))) %>%
  mutate(across(everything(), as.factor))
ggplot(patterns, aes(x = name, y = rownr, fill = value, type = pat)) +
  geom_tile(color = "black") +
  scale_fill_brewer(palette = "Set1") +
  facet_wrap(~pat) +
  theme_void() +
  theme(legend.position = "none") 
```

@@

- What other methods are available? General overview of approaches, see @audi18 @grun18. E.g., imputation of study level versus patient-level covariates, and one-stage imputation versus two-stage imputation methods.

-	Additional problem addressed in this paper: handling MNAR.

@@

- Case study options: metamisc::impact (IPD), GREAT (IPD).

@@

@@

- ADD: Section with more in-depth info on different data types and suitable methods. Add some equations, but not too much -> grow in complexity.

- Ideally we impute with random everything and heteroscedastic errors: most generic method (no worry about congeneiality) --> explain predictor matrix. Just mention that we should think about random intercept/slope, but just refer to other papers. This paper is just a software tutorial. --> Typically, at least random intercepts, but often random slopes as well. Refer to other papers for background, we'll focus just on the software implementation of the situations mentioned there. Sometimes there's little reason to assume some variable is affected by heterogeniety. We won't focus on congeneiality, that's too complex for 2-3 papers.

@@

- How to define the imputation model(s) in mice?

@@

@@

@@

- ADD: export mids objects to other packages like lme4 or coxme(?)

# Think about adding

- JOMO in mice --> on the side for now

@@

>>>>

---
documentclass: jss
author:
  - name: Hanne Oberman
    affiliation: Utrecht University
    address: |
      | Padualaan 14
      | 3584 CH Utrecht
    email: \email{h.i.oberman@uu.nl}
    url: https://hanneoberman.github.io/
  - name: Johanna Munoz Avila
    affiliation: University Medical Center Utrecht \AND
  - name: Valentijn de Jong
    affiliation: University Medical Center Utrecht 
  - name: Gerko Vink
    affiliation: Utrecht University \AND
  - name: Thomas Debray
    affiliation: University Medical Center Utrecht 
      # | Julius Center for Health Sciences and Primary Care, 
      # | University Medical Center Utrecht, Utrecht, The Netherlands
    #affiliation2:
title:
  formatted: "Imputation of Incomplete Multilevel Data with \\pkg{mice}"
  # If you use tex in the formatted title, also supply version without
  plain:     "Imputation of Incomplete Multilevel Data with mice"
  # For running headers, if needed
  short:     "\\pkg{mice}: Multilevel"
abstract: >
  Tutorial paper on imputing incomplete multilevel data with \pkg{mice}. Including methods for ignorable and non-ignorable missingness. 
keywords:
  # at least one keyword must be supplied
  formatted: [missing data, multilevel, clustering, "\\pkg{mice}", "\\proglang{R}"]
  plain:     [missing data, multilevel, clustering, mice, R]
preamble: >
  \usepackage{amsmath}
output: rticles::jss_article
bibliography: ../References/multilevelmice.bib
---

@@

- What is clustering/multilevel data? In this paper, we discuss grouped observations, not longitudinal data (within-patient clustering).  -> ADD: timeseries also in Discussion section.

- What do we mean by clustering? In the medical field: Clustering by studies (IPDMA), hospitals in registries, multi-center studies etc. In other fields: e.g. official stats clustering at country-level, or social sciences clustering at school-level (related to the sampling design).

- What is heterogeniety? I.e. variability within studies vs. variability between studies

- What does multilevel data look like? ADD: figure to show difference between patient-level datapoints vs cluster-level datapoints. Maybe also add different data frame formats (or just explain in text that there's long and wide formats).

- What methods are required to analyze multilevel data?	Add references, e.g. @hox17. At least explain difference random effects for intercept term, predictor effects, and/or variance residual error.

@@

- Why/where does missingness occur in multilevel data? I.e., not only patient-level but also cluster-level.

- How can we categorize this? Systematic vs sporadic missingness, see @resc13. ADD: visualization of systematic vs sporadic missingness. Within systematic we have always missing (same value per cluster) and non-measured variables (may differ per patient). TODO: adjust md pattern to match text. -> syst may vary or same for all patients (observations/participants).

```{r patterns, echo=FALSE}
# monotone vs non-monotone
patterns <- rbind(
  tibble(pat = rep("Systematic", 5), rownr = 5:1, X = rep(1, 5), Y = c(1,1,1,0, 0), Z = c(1, rep(0,4))) %>% pivot_longer(cols = c(X, Y, Z)),
  tibble(pat = rep("Sporadic", 5), rownr = 5:1, X = c(0, rep(1,4)), Y = c(1,1,1,0, 0), Z = c(1, rep(0,3), 1)) %>% pivot_longer(cols = c(X, Y, Z))) %>%
  mutate(across(everything(), as.factor))
ggplot(patterns, aes(x = name, y = rownr, fill = value, type = pat)) +
  geom_tile(color = "black") +
  scale_fill_brewer(palette = "Set1") +
  facet_wrap(~pat) +
  theme_void() +
  theme(legend.position = "none") 
```

-	ADD: missingness mech here

@@

- What types of multilevel methods are available? General overview of approaches, see @audi18 @grun18. E.g., imputation of study level versus patient-level covariates, and one-stage imputation versus two-stage imputation methods.

@@

- We focus on the workflow for conditional modeling (not JOMO) in `mice`. Refer to other packages: `mitml`, `miceadds`.

- Case study options: `metamisc::impact` (real data on traumatic brain injuries, IPD), `mice::popularity` (simulated data with MNAR/MAR mixture, schools). -> Check Gelman's data/NSRI data.

@@

## Case study

- Introduce case study and set scope of this tutorial: We're providing an overview of implementations. It's up-to the reader to decide which strategy suits their data. So we won't go into detail for the different methods (and equations). This paper is just a software tutorial. We'll keep it practical. -> ADD: some kind of help function that suggests a suitable pred matrix to the user, given a certain analysis model.

@@

- Which models will we discuss? We'll build the model to grow in complexity. The final model is the most complex but also the most versatile.

- Note on model complexity: Typically, we should at least use random intercepts, but often random slopes as well. Ideally we impute with random everything and heteroscedastic errors: most generic method (no worry about congeniality, but don't mention the term) ->  Refer to other papers for background, we'll focus just on the software implementation of the situations mentioned there. Sometimes there's little reason to assume some variable is affected by heterogeneity. -> Refer to Meng, Vincent, and a paper by Grund on congeniality and random slopes.

- Step 0: study as a predictor, AKA multilevel imputation for dummies. Doesn't work for syst missing.

@@

- How to define the imputation model(s) in `mice`?

@@

- Step 1: Intercept

- Step 2: Slope

- Step 3: Residuals

- Heckman model for MNAR

@@

- Analysis of scientific interest.

- Pooling using `mitml`.

@@

- ADD: export mids objects to other packages like `lme4` or `coxme`(?)

# Discussion

- JOMO in `mice` --> on the side for now

@@

- Timeseries: and polynomial relationship in the clustering.

### 2021-11-01 Multilevel Missing Data/Multilevel-Missing-Data.tex
Outline paper incl modeling steps

\author{
Hanne Oberman\\Utrecht University \And Johanna Munoz Avila\\University
Medical Center Utrecht \And Valentijn\\University Medical Center Utrecht
\AND Gerko Vink\\Utrecht University \And Thomas Debray\\University
Medical Center Utrecht
}
\title{Imputation of Incomplete Multilevel Data with \pkg{mice}}

\Plainauthor{Hanne Oberman, Johanna Munoz Avila, Valentijn, Gerko
Vink, Thomas Debray}
\Plaintitle{Imputation of Incomplete Multilevel Data with mice}
\Shorttitle{\pkg{mice}: Multilevel}

\Abstract{
Tutorial paper on imputing incomplete multilevel data with \pkg{mice}.
Including methods for ignorable and non-ignorable missingness. AKA
missing on so many levels.
}

\Keywords{missing data, multilevel, clustering, \proglang{R}}
\Plainkeywords{missing data, multilevel, clustering, R}

@@

\begin{itemize}
\item
  What is clustering/multilevel data? In this paper, we discuss grouped
  observations, not longitudinal data (within-patient clustering). What
  do we mean by clustering? In the medical field: Clustering by studies
  (IPDMA), hospitals in registries, multi-center studies etc. In other
  fields: e.g.~official stats clustering at country-level, or social
  sciences clustering at school-level (related to the sampling design).
\item
  What is heterogeniety (i.e.~variability within studies vs.~variability
  between studies)? ADD: Figure to show difference between patient-level
  datapoints vs cluster-level datapoints. And different data frame
  formats.
\item
  What methods are required to analyze multilevel data? Random effects
  for intercept term, predictor effects, and/or variance residual error.
  Add references, e.g. \citet{hox17}.
\end{itemize}

@@

\begin{itemize}
\item
  Why/where does missingness occur? I.e., not only patient-level but
  also cluster-level.
\item
  How can we categorize this? Systematic vs sporadic missingness, see
  \citet{resc13}. ADD: visualization of systematic vs sporadic
  missingness. Within systematic we have always missing (same value per
  cluster) and non-measured variables (may differ per patient). TODO:
  adjust md pattern to match text.
\end{itemize}

@@

\begin{itemize}
\item
  Why are standard (ad hoc) missing data methods not well suited?
\item
  What other methods are available? General overview of approaches, see
  \citet{audi18} \citet{grun18}. E.g., imputation of study level versus
  patient-level covariates, and one-stage imputation versus two-stage
  imputation methods.
\item
  Additional problem addressed in this paper: handling MNAR.
\end{itemize}

@@

\begin{itemize}
\item
  Provide practical guidelines with code snippets for imputation of
  incomplete multilevel data.
\item
  Case study options: metamisc::impact (IPD), GREAT (IPD).
\end{itemize}

@@

@@

\begin{itemize}
\item
  ADD: Section with more in-depth info on different data types and
  suitable methods. Add some equations, but not too much -\textgreater{}
  grow in complexity.
\item
  Ideally we impute with random everything and heteroscedastic errors:
  most generic method (no worry about congeneiality) --\textgreater{}
  explain predictor matrix. Just mention that we should think about
  random intercept/slope, but just refer to other papers. This paper is
  just a software tutorial. --\textgreater{} Typically, at least random
  intercepts, but often random slopes as well. Refer to other papers for
  background, we'll focus just on the software implementation of the
  situations mentioned there. Sometimes there's little reason to assume
  some variable is affected by heterogeniety. We won't focus on
  congeneiality, that's too complex for 2-3 papers.
\end{itemize}

@@

\begin{itemize}
\item
  How to define the imputation model(s) in mice?
\item
  What do the different implementations look like?
\end{itemize}

@@

\begin{itemize}
\item
  Pooling `regular' parameters vs more `exotic' parameters (SE of
  residual errors, or autocorreletion)
\item
  ADD: export mids objects to other packages like lme4 or coxme(?)
\end{itemize}

\hypertarget{think-about-adding}{%
\section{Think about adding}\label{think-about-adding}}

\begin{itemize}
\item
  JOMO in mice --\textgreater{} on the side for now
\item
  Additional levels of clustering
\end{itemize}

>>>>

\author{
Hanne Oberman\\Utrecht University \And Johanna Munoz Avila\\University
Medical Center Utrecht \AND Valentijn de Jong\\University Medical
Center Utrecht \And Gerko Vink\\Utrecht University \AND Thomas
Debray\\University Medical Center Utrecht
}
\title{Imputation of Incomplete Multilevel Data with \pkg{mice}}

\Plainauthor{Hanne Oberman, Johanna Munoz Avila, Valentijn de
Jong, Gerko Vink, Thomas Debray}
\Plaintitle{Imputation of Incomplete Multilevel Data with mice}
\Shorttitle{\pkg{mice}: Multilevel}

\Abstract{
Tutorial paper on imputing incomplete multilevel data with \pkg{mice}.
Including methods for ignorable and non-ignorable missingness.
}

\Keywords{missing
data, multilevel, clustering, \pkg{mice}, \proglang{R}}
\Plainkeywords{missing data, multilevel, clustering, mice, R}

@@

\begin{itemize}
\item
  What is clustering/multilevel data? In this paper, we discuss grouped
  observations, not longitudinal data (within-patient clustering).
  -\textgreater{} ADD: timeseries also in Discussion section.
\item
  What do we mean by clustering? In the medical field: Clustering by
  studies (IPDMA), hospitals in registries, multi-center studies etc. In
  other fields: e.g.~official stats clustering at country-level, or
  social sciences clustering at school-level (related to the sampling
  design).
\item
  What is heterogeniety? I.e. variability within studies vs.~variability
  between studies
\item
  What does multilevel data look like? ADD: figure to show difference
  between patient-level datapoints vs cluster-level datapoints. Maybe
  also add different data frame formats (or just explain in text that
  there's long and wide formats).
\item
  What methods are required to analyze multilevel data? Add references,
  e.g. \citet{hox17}. At least explain difference random effects for
  intercept term, predictor effects, and/or variance residual error.
\end{itemize}

@@

\begin{itemize}
\item
  Why/where does missingness occur in multilevel data? I.e., not only
  patient-level but also cluster-level.
\item
  How can we categorize this? Systematic vs sporadic missingness, see
  \citet{resc13}. ADD: visualization of systematic vs sporadic
  missingness. Within systematic we have always missing (same value per
  cluster) and non-measured variables (may differ per patient). TODO:
  adjust md pattern to match text. -\textgreater{} syst may vary or same
  for all patients (observations/participants).
\end{itemize}

@@

\begin{itemize}
\item
  ADD: missingness mech here
\item
  Why are standard (ad hoc) missing data methods not well suited?
\item
  What types of multilevel methods are available? General overview of
  approaches, see \citet{audi18} \citet{grun18}. E.g., imputation of
  study level versus patient-level covariates, and one-stage imputation
  versus two-stage imputation methods.
\end{itemize}

@@

\begin{itemize}
\item
  Provide practical guidelines with code snippets for imputation of
  incomplete multilevel data.
\item
  We focus on the workflow for conditional modeling (not JOMO) in
  \texttt{mice}. Refer to other packages: \texttt{mitml}.
\item
  Case study options: \texttt{metamisc::impact} (real data on traumatic
  brain injuries, IPD), \texttt{mice::popularity} (simulated data with
  MNAR/MAR mixture, schools). -\textgreater{} Check Gelman's data/NSRI
  data.
\end{itemize}

@@

\hypertarget{case-study}{%
\subsection{Case study}\label{case-study}}

\begin{itemize}
\tightlist
\item
  Introduce case study and set scope of this tutorial: We're providing
  an overview of implementations. It's up-to the reader to decide which
  strategy suits their data. So we won't go into detail for the
  different methods (and equations). This paper is just a software
  tutorial. We'll keep it practical. -\textgreater{} ADD: some kind of
  help function that suggests a suitable pred matrix to the user, given
  a certain analysis model.
\end{itemize}

@@

\begin{itemize}
\item
  Which models will we discuss? We'll build the model to grow in
  complexity. The final model is the most complex but also the most
  versatile.
\item
  Note on model complexity: Typically, we should at least use random
  intercepts, but often random slopes as well. Ideally we impute with
  random everything and heteroscedastic errors: most generic method (no
  worry about congeniality, but don't mention the term) -\textgreater{}
  Refer to other papers for background, we'll focus just on the software
  implementation of the situations mentioned there. Sometimes there's
  little reason to assume some variable is affected by heterogeneity.
  -\textgreater{} Refer to Meng, Vincent, and a paper by Grund on
  congeniality and random slopes.
\item
  Step 0: study as a predictor, AKA multilevel imputation for dummies.
  Doesn't work for syst missing.
\end{itemize}

@@

\begin{itemize}
\item
  How to define the imputation model(s) in \texttt{mice}?
\item
  What do the different implementations look like?
\item
  Step 1: Intercept
\item
  Step 2: Slope
\item
  Step 3: Residuals
\item
  Heckman model for MNAR
\end{itemize}

@@

\begin{itemize}
\item
  Analysis of scientific interest.
\item
  Pooling using \texttt{mitml}.
\item
  Pooling `regular' parameters vs more `exotic' parameters (SE of
  residual errors, or autocorreletion)
\item
  ADD: export mids objects to other packages like \texttt{lme4} or
  \texttt{coxme}(?)
\end{itemize}

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

\begin{itemize}
\item
  JOMO in mice --\textgreater{} on the side for now
\item
  Additional levels of clustering
\item
  Timeseries: and polynomial relationship in the clustering.
\end{itemize}
